---
title: "JSC370 Final Project"
author: "Yu Hau (Howard) Chen"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

<br>

The purpose of this webpage is to showcase some interactive visuals.

<br>

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}

library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)
library(readxl)
library("rvest")
library('xml2')
library(leaflet)

# INITALIZE CODE CHUNK OPTIONS
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = FALSE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
  class.source = "code-r")
```

```{css, echo = FALSE}
.code-r { /* Code block */
  font-size: 15px;
}

.code-r-small { /* Code block */
  font-size: 10px;
}
```

### Canada House Price Summary {.tabset}

The [Multiple Listing Service House Price Index](https://www.crea.ca/housing-market-stats/mls-home-price-index/hpi-tool/) dataset provies information of average home price and price index for various different real estates (Ex. Townhouse, apartment, etc), at different time point. We will use the dataset to plot the average home price at the year of 2022, as well as the percentage of growth of home price from January 2020 to January 2022 for different cities in Canada. We created a map, where you can see the locations of the cities. Moreover, by clicking on the markets on the map, you can see the percentage of growth and the average house price of the city. We also have created a scatterplot in representing Average House Price versus Percentage of Growth from Jan 2020 to Jan 2022. By hovering the mouse on the circles on the graph, we can see information including:

- Location
- Average House Price
- Average HPI (House Price Index)
- Percentage of Growth

From the graph, we observe that the cities that have both high price and high growth are Toronto and Hamilton. We also observe that inland cities such as Edmonton, Calgary, Winnipeg, and Saskatoon generally has relatively lower price and lower growth.

```{r, echo = FALSE, message = FALSE}
house_price_summary <- read_csv("data/canada_house_price_summary_2022.csv")
```

```{r, echo = FALSE}
scatter_p2 <- house_price_summary %>%
  plot_ly(x = ~Composite_Benchmark, y = ~Growth, type = 'scatter', 
          mode = "markers", color = ~Location, size = ~Growth,
          sizes = c(5, 70), marker = list(sizemode = "diameter", opacity = 0.5),
          hover_info = "text", text = ~paste(paste("Location: ", Location), 
                                             paste("Average House Price: ", Composite_Benchmark),
                                             paste("Average HPI: ", Composite_HPI),
                                             paste("Percentage of Growth: ", paste(round(Growth, digits = 3)*100, "%")),
                                             sep = "<br>")
          ) %>%
  layout(
         yaxis = list(title = "Percentage of Growth from Jan 2020 to Jan 2022"),
         xaxis = list(title = "Average House Price"),
         hovermodel = "compare")
```

```{r}
canada_content <- paste(sep = "<br/>",
  "<b>Canada</a></b>",
  "Percentage of Growth: 46.5%",
  "Average House Price: $825,800"
)
victoria_content <- paste(sep = "<br/>",
  "<b>Victoria</a></b>",
  "Percentage of Growth: 32.6%",
  "Average House Price: $920,400"
)
vancouver_content <- paste(sep = "<br/>",
  "<b>Vancouver</a></b>",
  "Percentage of Growth: 25.0%",
  "Average House Price: $1,255,200"
)
calgary_content <- paste(sep = "<br/>",
  "<b>Calgary</a></b>",
  "Percentage of Growth: 14.5%",
  "Average House Price: $458,800"
)
edmonton_content <- paste(sep = "<br/>",
  "<b>Edmonton</a></b>",
  "Percentage of Growth: 8.0%",
  "Average House Price: $339,600"
)
saskatoon_content <- paste(sep = "<br/>",
  "<b>Saskatoon</a></b>",
  "Percentage of Growth: 13.5%",
  "Average House Price: $328,600"
)
winnipeg_content <- paste(sep = "<br/>",
  "<b>Winnipeg</a></b>",
  "Percentage of Growth: 24.3%",
  "Average House Price: $331,300"
)
hamilton_content <- paste(sep = "<br/>",
  "<b>Hamilton</a></b>",
  "Percentage of Growth: 64.9%",
  "Average House Price: $1,058,000"
)
ottawa_content <- paste(sep = "<br/>",
  "<b>Ottawa</a></b>",
  "Percentage of Growth: 45.9%",
  "Average House Price: $689,700"
)
toronto_content <- paste(sep = "<br/>",
  "<b>Toronto</a></b>",
  "Percentage of Growth: 50.6%",
  "Average House Price: $1,259,900"
)
montreal_content <- paste(sep = "<br/>",
  "<b>Montreal</a></b>",
  "Total Growth Rate: 42.4%",
  "Average House Price: $530,100"
)
nb_content <- paste(sep = "<br/>",
  "Percentage of Growth: 55.7%",
  "Average House Price: $257,000"
)
nl_content <- paste(sep = "<br/>",
  "<b>Newfoundland and Labrador</a></b>",
  "Percentage of Growth: 20.5%",
  "Average House Price: $324,800"
)
```

```{r}
cb_pal <- colorNumeric(c("blue", "purple", "red"), 
                       domain = house_price_summary$Composite_Benchmark, na.color = NA)

map_p1 <- house_price_summary %>%
  leaflet() %>%
  addProviderTiles('OpenStreetMap') %>%
  addCircles(lat = ~lat, lng = ~lon, color = ~cb_pal(Composite_Benchmark),
             label = ~paste('$',formatC(house_price_summary$Composite_Benchmark, big.mark=',', format = 'f')), 
             opacity = 1, fillOpacity = 1) %>%
  addLegend('bottomleft', pal = cb_pal, values = house_price_summary$Composite_Benchmark, 
            title = "Average House Price", opacity = 1, na.label = NULL) %>%
  addMarkers(-106.3468, 56.1304, popup = ~canada_content) %>%
  addMarkers(-123.329773, 48.407326, popup = ~victoria_content) %>%
  addMarkers(-123.116226, 49.246292, popup = ~vancouver_content) %>%
  addMarkers(-114.066666, 51.0447, popup = ~calgary_content) %>%
  addMarkers(-113.4938, 53.5461, popup = ~edmonton_content) %>%
  addMarkers(-106.6702, 52.1579, popup = ~saskatoon_content) %>%
  addMarkers(-97.1384, 49.8951, popup = ~winnipeg_content) %>%
  addMarkers(-79.8711, 43.2557, popup = ~hamilton_content) %>%
  addMarkers(-75.6972, 45.4215, popup = ~ottawa_content) %>%
  addMarkers(-79.3832, 43.6532, popup = ~toronto_content) %>%
  addMarkers(-73.5673, 45.5017, popup = ~montreal_content) %>%
  addMarkers(-66.4619, 46.5653, popup = ~nb_content) %>%
  addMarkers(-57.6604, 53.1355, popup = ~nl_content) 
```

#### Map

```{r}
map_p1
```

#### Scatterplot

```{r}
scatter_p2
```

### Government of Canada Marketable Bonds Average Yield

The [Financial Market Statistics](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1010014501) dataset provides information about various financial market rates such as bank rate, interest rates, government bond yields, treasury bills, etc, on a weekly basis. For this interactive visualization, we will only be looking at the Government of Canada marketable bonds average yields. We created an interactive time series plot in Figure 2, with range slider that allows users to pan and zoom the X-axis, and range selector buttons that allow users to easily set the range of x-axis, that shows the yields of the bonds with different maturity from 1980 to 2022. We observe that there is an overall decreasing trend of bond yields. This caused by various factors, In the 1990s, inflation decline, which leads to increase in demand of government bonds, that further drives up the bond price and decreases bond yields. In the 2000s, investors and central bank has increases purchases of government bonds as it was an emerging market, and more recently, yields dropped close to zero due to quantitative easing and monetary policies.

```{r, echo=FALSE, message=FALSE}
fm_stats_df <- read_csv("Data/Financial Markets/10100145.csv")
fm_stats_df <- fm_stats_df %>% filter(grepl("^Government of Canada marketable bonds*", Rates))
fm_stats_df$REF_DATE <- as.Date(fm_stats_df$REF_DATE)
fm_stats_df_explore <- fm_stats_df %>% filter(REF_DATE >= "1980/01/01")
fm_stats_df_explore <- separate(fm_stats_df_explore, Rates, into = c("Type", "Length"), sep = ": ")
fm_stats_df_explore <- fm_stats_df_explore %>% drop_na(VALUE)
```

```{r, echo = FALSE}
fm_stats_df_explore %>%
  plot_ly(x = ~REF_DATE, y = ~VALUE, color = ~Length, type = "scatter",
          mode = "lines") %>%
  layout(
         yaxis = list(title = "Rate (%)"),
         xaxis = list(title = "Date",
      rangeselector = list(
        buttons = list(
          list(
            count = 3,
            label = "3 mo",
            step = "month",
            stepmode = "backward"),
          list(
            count = 6,
            label = "6 mo",
            step = "month",
            stepmode = "backward"),
          list(
            count = 1,
            label = "1 yr",
            step = "year",
            stepmode = "backward"),
          list(
            count = 1,
            label = "YTD",
            step = "year",
            stepmode = "todate"),
          list(step = "all"))),

      rangeslider = list(type = "date")),
         hovermodel = "compare")
```

### Mortgages in Arrears

From the website [Mortgages in Arrears](https://www.cmhc-schl.gc.ca/en/blog/2021/new-report-residential-mortgages-shows-30-year-low-arrears), we scraped the table “Mortgages in arrears (delinquent for 90 or more days) continued downward trend across all lender types”. The table contains information about mortgage in arrears across different lender types from 2020 Q1 to 2021 Q1. An interactive grouped bar chart in Figure 3 was created, displaying the percentage of mortgage in arrears for different mortgage lenders by time. We observe that although the mortgage in arrears increases from 2020 Q1 to 2020 Q2, it has started to decrease after that. The reason of this is because of quantitative easing. 

```{r, echo = FALSE}
url <- "https://www.cmhc-schl.gc.ca/en/blog/2021/new-report-residential-mortgages-shows-30-year-low-arrears"
mortgage_arrears <- read_html(url)
table <- xml_find_all(mortgage_arrears, xpath='/html/body/div[3]/main/section/div[9]/div/figure/div/div[2]/div/table')
table <- html_table(table)
mortgage_arrears_df <- data.frame(t(table[[1]]))
colnames(mortgage_arrears_df) <- mortgage_arrears_df[1,]
mortgage_arrears_df <- mortgage_arrears_df[-1, ]
mortgage_arrears_df <- cbind(`Mortgage Lenders` = rownames(mortgage_arrears_df), mortgage_arrears_df)
rownames(mortgage_arrears_df) <- 1:nrow(mortgage_arrears_df)
mortgage_arrears_df$`Mortgage Lenders`[3] <- "Other non-bank lenders"
mortgage_arrears_tidy <- mortgage_arrears_df %>% 
                         tidyr::pivot_longer(
                           cols = !`Mortgage Lenders`, 
                           names_to = "Quarter", 
                           values_to = "Arrears")
mortgage_arrears_tidy$Arrears <- as.double(substr(mortgage_arrears_tidy$Arrears, 1, 4), digits=3)
```

```{r, echo = FALSE}
mortgage_arrears_tidy %>% 
  plot_ly(x = ~`Mortgage Lenders`, y = ~Arrears, type = 'bar', color = ~Quarter) %>%
  layout(
         yaxis = list(title = "Mortgage in Arrears (%)"),
         xaxis = list(title = "Mortage Lenders"),
         hovermodel = "compare")
```

Note that MIEs stands for Mortgage Investment Entities.


### CPI during COVID-19 Pandemic

Real estate is known to be an investment product that performs well in high-inflation environments, which leads to a higher demand. The interactive visualizations below show that trend of CPI from January 2020 to January 2021, as well as breaking down the CPI for major categories such as food, shelter, clothing and footwear, health and personal care, energy, and services. By hovering mouse onto the markets, you are able to see the exact number of CPI at each time point. We observe that the energy sector has the most price growth, by roughly 50% from April 2020 to Jan 2020. Overall, CPI has raised significantly since 2021, and the not only the trend is increasing, the momentum is increasing as well.

```{r, message=FALSE}
cpi_df <- read_csv("data/cpi.csv")
```

```{r}
all_items <- cpi_df %>% 
  filter(`Products and product groups` == "All-items") %>%
  plot_ly(x = ~REF_DATE, y = ~VALUE, type = "scatter",
          mode = "lines+markers") 
food <- cpi_df %>% 
  filter(`Products and product groups` == "Food") %>%
  plot_ly(x = ~REF_DATE, y = ~VALUE, type = "scatter",
          mode = "lines+markers") 
shelter <- cpi_df %>% 
  filter(`Products and product groups` == "Shelter") %>%
  plot_ly(x = ~REF_DATE, y = ~VALUE, type = "scatter",
          mode = "lines+markers")
clothing <- cpi_df %>% 
  filter(`Products and product groups` == "Clothing and footwear") %>%
  plot_ly(x = ~REF_DATE, y = ~VALUE, type = "scatter",
          mode = "lines+markers")
health <- cpi_df %>% 
  filter(`Products and product groups` == "Health and personal care") %>%
  plot_ly(x = ~REF_DATE, y = ~VALUE, type = "scatter",
          mode = "lines+markers")
energy <- cpi_df %>% 
  filter(`Products and product groups` == "Energy") %>%
  plot_ly(x = ~REF_DATE, y = ~VALUE, type = "scatter",
          mode = "lines+markers")
services <- cpi_df %>% 
  filter(`Products and product groups` == "Services") %>%
  plot_ly(x = ~REF_DATE, y = ~VALUE, type = "scatter",
          mode = "lines+markers")

s1 <- subplot(food, shelter, clothing, health, energy, services, nrows = 3, titleY = FALSE, titleX = FALSE, shareX = TRUE, margin = 0.05)
annotations = list( 
  list( 
    x = 0.22,  
    y = 1.0,  
    text = "Food",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ),  
  list( 
    x = 0.77,  
    y = 1,  
    text = "Shelter",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ),  
  list( 
    x = 0.22,  
    y = 0.65,  
    text = "Clothing and Footwear",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ),
  list( 
    x = 0.77,  
    y = 0.65,  
    text = "Health and Personal Care",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ),
  list( 
    x = 0.22,  
    y = 0.30,  
    text = "Energy",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ),
  list( 
    x = 0.77,  
    y = 0.30,  
    text = "Services",  
    xref = "paper",  
    yref = "paper",  
    xanchor = "center",  
    yanchor = "bottom",  
    showarrow = FALSE 
  ))

s1 <- s1 %>% layout(annotations = annotations, showlegend = FALSE)
fig <- subplot(all_items, s1, nrows = 2, margin = 0.05, heights = c(0.25,0.75)) %>% layout(title = "CPI during COVID-19 Pandemic")
fig
```



<br>
<br>